Efficiently stable presentations from error-correcting codes

### 2023-09-18 efficient_stability.tex
finished introduction first pass

\paragraph{Groups stability.}
This observation, which at first may seem quite anodyne or even forced, opens a universe of questions. The problem of relating approximate representations of a group to exact representations of it is termed \emph{stability} in group theory, and has a long history. There are of course many flavors of the problem, depending on how one defines closeness (should it be the operator norm or the Hilbert-Schmidt norm? Should closeness hold for every relation, or every pair of group elements, or is it sufficient that it holds on average? Etc.) We will review some relevant results in this area below. For now, we can mention Voiculescu's famous counter-example~\cite{voiculescu1983asymptotically} about approximately commuting unitaries, which was motivated by a question of Halmost on pairs of approximately commuting Hermitian operators~\cite{halmos1976some}. Using the terminology of stability, Voiculeschu showed that the presentation $\Z^2 = \langle x,y:[x,y]=e\rangle$ is \emph{not} stable with respect to the operator norm. However, much more recently Glebsky~\cite{glebsky2010almost} showed that the same presentation \emph{is} stable with respect to the normalized Hilbert-Schmidt norm.

@@

Our main results are presentations of $\Z_2^k$, and of slightly more complex $2$-groups built on them (e.g. the \emph{Pauli}, or \emph{Weyl-Heisenberg}, group ubiquitous in quantum information theory), that are \emph{efficiently stable}: the size of the presentation is quasi-polynomial, as opposed to exponential, in $k$, and yet the modulus of stability only depends poly-logarithmically on $k$. These presentations are constructed from specific error-correcting codes, the Reed-Muller codes based on polynomials, which are known to have good local testability properties~\cite{}.

@@

We do not know of any other presentation of $\Z_2^k$, which is arguably the simplest example one could think of, that is stable with similar parameters as the one stated in the theorem. It would be very interesting to discover different such presentations, built from locally testable error-correcting codes or not, and for this group or other groups.

\paragraph{Nonlocal games in quantum information theory.}

To motivate the focus on efficient stability we now sketch a connection between our results and problems in quantum complexity theory and in particular the theory of nonlocal games that motivate us. A \emph{nonlocal game} $\game$ is specified by the following date: finite question and answer sets $\mX$ and $\mA$ respectively, a distribution $\mu$ on $\mX\times \mX$, and a decision predicate $D:\mX\times \mX\times \mA\times \mA\to \{0,1\}$. The interpretation of $G=(\mX,\mu,\mA,D)$ as a game is as follows. A ``referee'' is imagined to sample a pair of ``questions'' $(x,y)\sim \mu$. Each question is sent to a different player, who is tasked with responding with an answer $a,b\in \mA$ respectively. Finally, the referee decides that the players win the game if and only if $D(x,y,a,b)=1$. Interestingly, the maximum success probability of the players in this game, probability is over the referee's choice of questions and any randomness in the player's strategy, depends on the set of strategies that is allowed. In particular, and without going into details, this maximum can be different depending on whether only \emph{classical} strategies are allowed (where each player answers according to a function $f:\mX\to\mA$ of their choice), or \emph{quantum} strategies are allowed, which allow some additional form of coordination between the players in the form of \emph{quantum entanglement}. Understanding when there is a gap between the two values, and how large the gap can be, is of great interest in the foundations of quantum mechanics. To study this question one is drawn to investigate the structure of optimal quantum strategies in a game, and how to design games that enforce a specific structure --- for example, forcing as much ``non-classicality'' in the strategy as possible. (And beyond that, many applications to e.g.\ verification which we do not survey in detail.)

A class of games that has been successful as LCS. These are defined by a system of equations. We now see the connection with presentations of the form $G(h)$. The stability results enable conclusions about approximately optimal strategies. In particular, the results about the Pauli group lead to qubit and dimension tests. These are an ingredient in MIP*=RE. Recover PBT

\paragraph{Related works on stability.}

\paragraph{Outline.}

@@

\section{Efficient stability}

@@

\section{Presentations from codes}

@@

\section{Testing entanglement}

@@

\subsection{Nonlocal games}

@@

\subsection{The Pauli braiding test}

>>>>

\paragraph{Group stability.}
This observation, which at first may seem quite anodyne or even forced, opens a universe of questions. The problem of relating approximate representations of a group to exact representations of it is termed \emph{stability} in group theory, and has a long history. There are of course many flavors of the problem, depending on how one defines closeness (should it be the operator norm or the Hilbert-Schmidt norm? Should closeness hold for every relation, or every pair of group elements, or is it sufficient that it holds on average? Etc.) We will review some relevant results in this area below. For now, we can mention Voiculescu's famous counter-example~\cite{voiculescu1983asymptotically} about approximately commuting unitaries, which was motivated by a question of Halmost on pairs of approximately commuting Hermitian operators~\cite{halmos1976some}. Using the terminology of stability, Voiculeschu showed that the presentation $\Z^2 = \langle x,y:[x,y]=e\rangle$ is \emph{not} stable with respect to the operator norm. However, much more recently Glebsky~\cite{glebsky2010almost} showed that the same presentation \emph{is} stable with respect to the normalized Hilbert-Schmidt norm.

@@

Our main results are presentations of $\Z_2^k$, and of slightly more complex $2$-groups built on them (e.g. the \emph{Pauli}, or \emph{Weyl-Heisenberg}, group ubiquitous in quantum information theory), that are \emph{efficiently stable}: the size of the presentation is quasi-polynomial, as opposed to exponential, in $k$, and yet the modulus of stability only depends poly-logarithmically on $k$. These presentations are constructed from specific error-correcting codes, the Reed-Muller codes based on polynomials, which are known to have good local testability properties~\cite{}. Our main result on group stability can be stated as follows (see Theorem~\ref{thm:z2-stab} for the precise statement).

@@

We do not know of any other presentation of $\Z_2^k$, arguably the simplest example one could think of, that is stable with similar parameters as the ones stated in the theorem. It would be very interesting to discover different such presentations, built from locally testable error-correcting codes or not, for this group or others.

\paragraph{Nonlocal games in quantum information theory.}
To motivate our focus on \emph{efficient} stability we now sketch a connection between our results and problems in quantum complexity theory and in particular the theory of nonlocal games that motivate us. As a result we will recover a key technical result used in the proof of the recent MIP$^*=$RE~\cite{ji2020mip}.

 A \emph{nonlocal game} $\game$ is specified by the following data: finite question and answer sets $\mX$ and $\mA$ respectively, a distribution $\mu$ on $\mX\times \mX$, and a decision predicate $D:\mX\times \mX\times \mA\times \mA\to \{0,1\}$ (see Section~\ref{sec:nl-games} for details). The interpretation of $G=(\mX,\mu,\mA,D)$ as a game is as follows. A ``referee'' is imagined to sample a pair of ``questions'' $(x,y)\sim \mu$. Each question is sent to a different player, who is tasked with responding with an answer $a,b\in \mA$ respectively. Finally, the referee decides that the players win the game if and only if $D(x,y,a,b)=1$. Interestingly, the maximum success probability of the players in this game, where the probability is over the referee's choice of questions and any randomness in the player's strategy, and the maximum is taken over all allowed strategies, depends on whether one relies on ``classical'' or ``quantum'' interpretations of the game to determine appropriate mathematical formalization of the set of strategies that the players may employ. While a classical viewpoint naturally models a strategy as a pair of functions $f,g:\mX\to\mA$, one for each player, quantum mechanics invites one to consider a broader set of strategies in which an additional form of coordination between the players is allowed in the form of \emph{shared quantum entanglement}. Understanding when there is a gap between the resulting maxima, and how large this gap can be, is of great interest in the foundations of quantum mechanics. To study this question one is drawn to investigate the structure of optimal quantum strategies in a game, and how to design games that enforce a specific structure --- informally, forcing as much ``non-classicality'' in winning strategies as possible. Beyond their foundational appeal, the theory of nonlocal games has had a very large impact in quantum cryptography (such as the analysis of device-independent quantum key distribution protocols~\cite{vazirani2019fully,arnon2019simple}) and quantum complexity, in particular the theory of multiprover interactive proof systems~\cite{cleve2004consequences}.

For concreteness let us focus on a class of games called \emph{linear constraint system} (LCS) games. These games were introduced in~\cite{cleve2014characterization} and their study plays a central role in the celebrated result by Slofstra showing non-closure of the ``quantum set''~\cite{slofstra2019set}. A linear constraint system game is parametrized by a matrix $h\in \F^{m\times n}$. In the game $\game_h$, the referee selects a pair $(i,j)\in \{1,\ldots,m\}\times\{1,\ldots,n\}$ uniformly at random conditioned on $h_{ij}=1$. They send $i$ to the first player and $j$ to the second. The first player returns values in $\F$ for \emph{each} $j'$ such that $h_{ij'}\neq 0$, whereas the second players returns a single value in $\F$. The players win if the first player's answers satisfy the parity constraint, and the players' answers are consistent (the first player's answer associated with index $j$ matches the second player's answer).

Now we see that to each matrix $h$ we have associated a group $G(h)$, and a game $\game_h$. Moreover, and quite interestingly, there is a one-to-one correspondence between representations of $G(h)$ and \emph{perfect strategies} in $\game_h$, i.e.\ strategies that have success probability $1$ in the game.\footnote{To be precise, the group here is $G(h)$ to which have been added pairwise commutation conditions $[x_j,x_{j'}]=e$ whenever there is an $i$ such that $h_{ij}$ and $h_{ij'}$ are both nonzero. This correspondence was established for finite-dimensional strategies, and finite-dimensional representations, in~\cite{cleve2014characterization}. Extensions to infinite-dimensional strategies and representations have appeared in~\cite{cleve2017perfect}. See also e.g.~\cite{kim2018synchronous} for generalizations to the broader class of \emph{synchronous games}.}
This correspondence enables one to ``embed'' group representations into quantum strategies, thereby forcing them to demonstrate a high level of complexity; an approach that was at the heart of~\cite{kim2018synchronous}. Going further, for applications one is often required to understand not only optimal but also near-optimal strategies, whose success probability is e.g.\ $1-\eps$. The same correspondence associates to such strategies approximate homomorphisms of $G(h)$~\cite{slofstra2018entanglement}. Here once can see that measuring closeness on average over the choice of relation is a natural choice, which is all but forced by the definition of a game, where the questions are selected according to some pre-specified distribution.

To summarize, approximate stability results for $G(h)$ enable one to obtain structural results about near-optimal strategies in $\game_h$ --- such a result is known as a \emph{rigidity} result in quantum information. A rigidity result about a game called the Pauli braiding test is at the heart of the proof of MIP$^*$=RE, which led to a disproof of Connes' Embedding Problem. The analysis of this test requires an efficient stability result for the Weyl-Heisenberg group. Informally, the reason that the stability result needs to be for an efficient presentation of the Weyl-Heisenberg group, as opposed to e.g.\ the multiplication table presentation, is because to obtain the final result it is necessary that the ``complexity'' of the test, or game, is smaller than the ``complexity'' of the object, or group being tested. Of course here we are loose about what we mean by ``complexity,'' and refer to~\cite{ji2021mip,vidickmip} for high-level explanations. The Pauli braiding test is described and analyzed in Section~\ref{sec:pbt}. We end by mentioning an open question: if one was able to obtain $|S_k|,|R_k|=\poly(k)$ while also having $\delta(\eps)=\poly(\eps)$ in the informal theorem stated above, then this result would likely, through the connection we just described, have consequences for the efficient verification of quantum computations in the framework of interactive proof systems---see e.g.~\cite{coladangelo2019verifier,natarajan2023bounding} for a sample of known results in this direction.

\paragraph{Related works on stability.} The general question of group stability was first formulated by Ulam~\cite{ulam1960collection}, and later studied by Kazhdan for the case of the operator norm~\cite{kazhdan1982e}.

Stability of finite groups, with respect to the multiplication table representation, is shown in~\cite{gowers2017inverse}. 
One can also consider representations in permutations, see~\cite{glebsky2009almost,becker2022stability}.

\tnote{more? open questions?}

\paragraph{Outline.}
We start in Section~\ref{sec:efficient} by giving a precise definition of stability that we work with, and give examples to motivate and illustrate the definition. In Section~\ref{sec:pres-codes} we give a general method for constructing presentations from codes, and apply the method to the spacial case of the Reed-Muller code. This leads in Section~\ref{sec:eff-z2k} to the statement and proof of our main result, an efficiently stable presentation for $\Z_2^k$. Finally in Section~\ref{sec:quantum} we elaborate on the connection with the theory of nonlocal games and detail our applications to this area.

@@

\section{Efficient stability}
\label{sec:efficient}

@@

\section{Presentations from codes}
\label{sec:pres-codes}

@@

\section{Testing entanglement}
\label{sec:quantum}

@@

\subsection{Nonlocal games}
\label{sec:nl-games}

@@

\subsection{The Pauli braiding test}
\label{sec:pbt}
